Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-08-08 bmc_template/manuscript.tex
acknowledgements

%%% Load packages
%\usepackage{amsthm,amsmath}
\usepackage{amsmath}
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{graphicx}
% \def\includegraphic{}
% \def\includegraphics{}

@@

\titlepage
\

@@

@@

\begin{artnotes}
 \note{Sample of title note}     % note to the article
 \note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

@@

\parttitle{Conclusion} %if any
Incorportating temporal heterogeneity reduce the selected number of behavioural states closer 
to a biologically interpretable level, improved goodness of fit and predictability. Our 
suggest that incorporating previously neglected structure in statistical models can
allow more accurate assessment of appropriate model complexity.

@@

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
van de Kerk et al.\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).

van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by van de Kerk et al.\cite{kerk2015hidden}.
Here we focus on the four cats with the most data (all with approximately 10,000-15,000
observations: see Table~1), 
in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets.  
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\cite{turchin1998quantitative,kerk2015hidden}.

van de Kerk et al.\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\cite{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by van de Kerk et al.\cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not given), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which corresponds
to making precisely this assumption.

van de Kerk et al.\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure 2).
(Note that most movement analyses, including van de Kerk et al.
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

van de Kerk et al.\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
three to six states,
suggesting that the six-state model was
favoured statistically; however,
the authors used three-state models in most
of their analyses for ease of biological interpretation.
We follow van de Kerk et al.\cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $ \mathbf Y = \obs_{1},..., \obs_{T}$ 
and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model
parameters \boldmath{$\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:

@@

We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of Morales et al.\cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.

@@

As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas HMMs with temporal heterogenity will increase quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

@@

The BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2)

@@

Models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

@@

The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \cite{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to van de Kerk et al.\cite{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state.

@@

On a broad spectrum, it really depends on what kind of question 
that is being answered. On one side of the spectrum, if the goal is 
to identify states, it might be sufficient to use a 
simple/traditional HMM model and pre-specify the number of states and, post hoc, 
match Viterbi-based states estimates with environmental variation 
\cite{fryxell_multiple_2008}. On the other side of the 
spectrum, if the goal of interest is to make predictions (out of sample), 
it might be better to fit a covariate-dependent model so that we can explicitly
model the switching process. In that case, fitting a covariate-dependent model is
better for out of sample prediction because Viterbi can only estimate state 
occupancy if observed movements are available (within sample preditions). 
Finally, if we want to estimate the number of states, BIC is not necessarily 
good for estimation of number of states \cite{biernacki2000assessing}, 
but it can be useful as an approximate upper limit estimate.

Incorporating temporal heterogenity in animal movement is one step in
the right direction, but much remains to be done. 
Our model neglects other predictors, such as habitat type or
location with respect to environmental features such as roads,
that can potentially improve goodness of fit and predictions and 
further reduce the estimated number of states. While adding
more covariates is in principle straightforward using existing
frameworks, including all possible biological complexities in 
a HMM with state-dependent transitions may rapidly become intractable
in terms of both computational time and complexity of choosing
among possible reduced models and numbers of states.
Better diagnostic procedures and tests are needed:
these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
and, more importantly, localize fitting problems to particular aspects of the
data so that models can be constructed without needing to
include all possible features of interest.

@@

\section*{Declarations}

@@

\section*{Acknowledgements}
  Florida Wildlife and Fisheries?

\section*{Funding}
 NSERC?

\section{Data accessibility}
Hourly step lengths and turning angles of male and female Florida panthers available at: http://ufdc.ufl.edu//IR00004241/00001.

\section*{Author's contributions}
    Equally contributed.

>>>>

%%% Load packages
%\usepackage{amsthm,amsmath}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{url}
\linenumbers
\RequirePackage[sort]{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{graphicx}
% \def\includegraphic{}
% \def\includegraphics{}

@@

%\setcounter{page}{0}  %% ugh, article template assumes starting on odd page!
% \newpage

@@

%% TO DO
% author references (maybe fix programmatically?)
% acknowledgments

@@

\begin{artnotes}
% \note{Sample of title note}     % note to the article
% \note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

@@

\parttitle{Conclusion} %if any
Incorporating temporal heterogeneity reduce the selected number of behavioural states closer 
to a biologically interpretable level, improved goodness of fit and predictability. Our 
suggest that incorporating previously neglected structure in statistical models can
allow more accurate assessment of appropriate model complexity.

@@

\doublespacing

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
van de Kerk et al. \cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).

van de Kerk et al. \cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by van de Kerk et al. \cite{kerk2015hidden}.
Here we focus on the four cats with the most data (all with approximately 10,000-15,000
observations: see Table~1), 
in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets.  
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\cite{turchin1998quantitative,kerk2015hidden}.

van de Kerk et al. \cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\cite{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by van de Kerk et al. \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not given), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which corresponds
to making precisely this assumption.

van de Kerk et al. \cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure 2).
(Note that most movement analyses, including van de Kerk et al.
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

van de Kerk et al. \cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
three to six states,
suggesting that the six-state model was
favoured statistically; however,
the authors used three-state models in most
of their analyses for ease of biological interpretation.
We follow van de Kerk et al. \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $ \mathbf Y = \obs_{1},..., \obs_{T}$ 
and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model
parameters {\boldmath $\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:

@@

We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of Morales et al. \cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.

@@

As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogeneity will increase linearly, whereas HMMs with temporal heterogeneity will increase quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

@@

The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2)

@@

Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

@@

The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \cite{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state.

@@

On a broad spectrum, it really depends on what kind of question 
that is being answered. On one side of the spectrum, if the goal is 
to identify states, it might be sufficient to use a 
simple/traditional HMM model and pre-specify the number of states and, post hoc, 
match Viterbi-based states estimates with environmental variation 
\cite{fryxell_multiple_2008}. On the other side of the 
spectrum, if the goal of interest is to make predictions (out of sample), 
it might be better to fit a covariate-dependent model so that we can explicitly
model the switching process. In that case, fitting a covariate-dependent model is
better for out of sample prediction because Viterbi can only estimate state 
occupancy if observed movements are available (within sample predictions). 
Finally, if we want to estimate the number of states, BIC is not necessarily 
good for estimation of number of states \cite{biernacki2000assessing}, 
but it can be useful as an approximate upper limit estimate.

Incorporating temporal heterogeneity in animal movement is one step in
the right direction, but much remains to be done. 
Our model neglects other predictors, such as habitat type or
location with respect to environmental features such as roads,
that can potentially improve goodness of fit and predictions and 
further reduce the estimated number of states. While adding
more covariates is in principle straightforward using existing
frameworks, including all possible biological complexities in 
a HMM with state-dependent transitions may rapidly become intractable
in terms of both computational time and complexity of choosing
among possible reduced models and numbers of states.
Better diagnostic procedures and tests are needed:
these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
and, more importantly, localize fitting problems to particular aspects of the
data so that models can be constructed without needing to
include all possible features of interest.

@@

@@

\section*{Acknowledgements}

We would like to thank Madelon van de Kerk, Madan Oli, and David Onorato for their previous work on Florida panthers. We also would like to thank McMaster University, Florida Fish and Wildlife Conservation Commission and many individuals for data collection and fieldwork. Lastly, to Madelon van de Kerk for making the data available at the Institutional Repository at the University of Florida (IR@UF).

\section*{Ethics approval}

All data used are secondary, drawn from an existing institutional
data repository.

\section*{Consent for publication}

Not applicable.

\section*{Funding}

This study was funded by NSERC Discovery Grant 386590-2010 to BMB.

\section{Data accessibility}
Hourly step lengths and turning angles of male and female Florida panthers available at: \url{http://ufdc.ufl.edu//IR00004241/00001}.

\section*{Authors' contributions}
ML designed analyses and simulations; ran analyses and simulations; and co-wrote the text of the paper.  BMB designed analyses and simulations and co-wrote the text of the paper.
